## Dulles Flights Data

The `flights_df` data frame is loaded below and consists of 33,433 flights from IAD (Dulles International) in 2016. The rows in this data frame represent a single flight with all of the associated features that are displayed in the table below.


```{r}

library(tidyverse)
library(ggplot2)
library(hrbrthemes)
flights_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/dulles_flights.rds'))

```


### Raw Data

```{r}

flights_df

```




## Question 1



**Question**:
Which airlines operate the most from Dulles International Airport and which destination city have more number of flights?

**Answer**:
The data indicates that United airlines has a fleet of 20,653 whose flights travel to different destinations followed by American Airlines having a fleet of 2597 flights. An intriguing observation has been observed where almost out of the 33,433 flights, 61.4% of the flights were operated by United airlines in the year 2016.

Among the 33,433 flights, 4034 flights flew to San Francisco making it as the busiest destination airport city where flights from IAD flew in 2016 and cities like Cleveland, Charlotte, Kansas City, Louisville and St.Louis have only 1 flight operating from the Dulles airport in the entire year. In conclusion, there are more number of flights coming from San Francisco.


To add additional R code chunks for your work, select `Insert` then `R` from the top of this notebook file.

```{r}
count(flights_df,airline)
#Airline carrier frequency
ggplot(flights_df,aes(airline))+geom_bar(position = position_dodge(width = 0.5),fill="light green")+ 
  geom_text(aes(label=..count..),stat="count",vjust = 1.2, colour = "black")+labs(title = "Airline carrier Frequency",position="center")+xlab("Airline")+ylab("Frequency")+theme(plot.title = element_text(hjust = 0.5),axis.text.x = element_text(size=8,angle=90))


#Flights operating to different cities
flights_df%>%
  arrange(desc(dest_airport_city))%>%
  count(dest_airport_city)
```



## Question 2



**Question**:
Is there a certain airline that has a record of number of delays in flight?


**Answer**:
Yes, United airlines has 3006 flights with arrival delay of more than 15 mins as per FAA. But there is one such flight of Delta airlines which has an arrival delay of 914 mins and departure delay of 911 mins, making it the most delayed flight in the entire year.In other words, Alaska airlines have less departure and arrival delay.The majority of flights departing on time arrive late for all the airlines. In conclusion, only by leaving early you can expect to arrive on time, and a reasonable cutoff for this seems to be around five minutes before scheduled takeoff.



```{r}

#Airlines having arrival delay of 15mins are considered as late (FAA)
flights_df%>%
  select(airline,arrival_delay,dep_delay)%>%
  filter(airline =="United"|airline =="American"|airline =="Delta"|airline =="Southwest"|airline =="JetBlue"|airline =="Virgin America"|airline =="Frontier"|airline =="ExpressJet"|airline =="SkyWest"|airline =="Alaska",arrival_delay>15)%>%
  arrange(desc(arrival_delay))%>%
  count(airline)

#Arrival and departure delay of all airlines
flights_delay<-ggplot(flights_df, aes(x= dep_delay, y=arrival_delay,color=airline)) +xlim(-30, 1000) + geom_point()+
  labs(title = "Airlines that have record number of delays")+
  xlab("Departure delay (mins)")+
  ylab("Arrival Delay (mins)")+theme(plot.title = element_text(hjust = 0.5))

ggplotly(flights_delay)

```



## Question 3



**Question**:
Is there any relation between the number of flights and the average_delay?


**Answer**:
From the scatter plot, we can see that there is much greater variation in average delay when there are few flights. For instance from the data, 125 flights having the tail number as N63890 have an average delay of -11.64 mins meaning these flights arrive almost 12 mins early to the airport. In conclusion, we can say that more number of flights have less arrival delays.


```{r}
delays<-flights_df%>%
group_by(tail_num)%>%
  summarize(delay=mean(arrival_delay,na.rm=TRUE),
            n=n())

delays

ggplot(delays,aes(x=n,y=delay))+geom_point(alpha=0.1)+ggtitle("Number of Flights Vs Average Delay")+xlab("Number of flights")+ylab("Average delay (mins)")+theme(plot.title = element_text(hjust = 0.5))

```

## Question 4


**Question**:
 Which destinations have the most airlines operating from Dulles airport?


**Answer**:
From the heatmap/tile chart depicted below, we can see that Atlanta has 5 different airlines i.e. United, Delta, Southwest, Frontier and ExpressJet operating from Dulles Airport. We can also see that United airlines travels to all the cities and operate through out the year. Also there are certain places like Charleston, St.Louis, Kansas City, Louisville, Cleveland, Eagle, Charlotte where only United Airlines operate.

In conclusion, United airlines is the most busiest airlines which travels to almost every city and operates each month of the year.



```{r}
flights_df %>%
    group_by(dest_airport_city) %>%
    summarize(airlines = n_distinct(airline)) %>%
    arrange(desc(airlines))
    ggplot(flights_df, aes(x=month, y=dest_airport_city, 
                           fill=airline,width=.7,height=.7))+
    geom_tile()+theme(axis.text.x = element_text( color="#993333", size=8,angle=90),
    axis.text.y = element_text( color="#993333",size=8))+
    ggtitle("Number of airlines operating to Destination Airport Cities")+
    xlab("Month")+ylab("Destination airport City")+
    theme(plot.title =element_text(hjust = 0.5))
    
    
```




## Question 5



**Question**:

What percentage of flights are delayed by more than one hour?


**Answer**:
The data indicates that  Frontier airlines have always had a departure delay of more than 1 hour. And during the month of July, 27% of the Frontier airlines have had a departure delay of more than 1 hour (July) which is the highest in the year 2016 when compared to other airlines. During the months of March, April, May, July and August, ExpressJet had no departure delay more than one hour.



```{r}
avg_dep_delay_df<-flights_df%>%
  group_by(month_numeric, airline) %>%
  summarize(avg_dep_delay = mean(dep_delay>60)*100,"%") %>%
  arrange(desc(avg_dep_delay))

avg_dep_delay_df

  ggplot(avg_dep_delay_df,mapping = aes(x = month_numeric, y = avg_dep_delay, color = airline)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = 1:12) +
  xlab("Month") +
  ylab("Average Departure Delay Time (percentage)") +
  ggtitle("Airlines having departure delay time more than 1 hour")+
  theme(plot.title =element_text(hjust = 0.5))
  

```


## Question 6



**Question**:
From which cities do flights arrive late to the Dulles International Airport?


**Answer**:
Flights from cities like Minneapolis-St.Paul,Dallas Fort-Worth and Detroit arrive late when compared to other cities. Flights from these cities have an arrival delay of more than 500 minutes. Whereas there are certain cities like San Francisco, Los Angeles, San Diego where certain flights arrive early even before their scheduled arrival time.



```{r}

#Arrival delay of flights from cities
flights_df %>%
  group_by(dest_airport_name,arrival_delay)%>%
  ggplot( aes(x = dest_airport_city, y = arrival_delay)) +
  geom_boxplot()+theme(axis.text.x = element_text(size=8,angle=90))+
  xlab("City Name") +
  ylab("Arrival delay of the flight \n in minutes") +
  ggtitle("Arrival delay of flights from cities")+
  theme(plot.title =element_text(hjust = 0.5))
```


## Question 7



**Question**:
Is there any certain season of the year where flights travel to distant cities having a departure delay of more than 15mins?


**Answer**:
According to the data, during the summer season, flights travel to distant places and among those flights United Airlines travels the most during each season with a departure delay of more than 15 mins. And during the Fall season, less number of flights travel to long distances. In simpler words, September, October and November are the months where airline carriers travel to less distant places.



```{r}
flights_df<-flights_df %>%
 mutate(season = ifelse(month_numeric %in% 9:11, "Fall",
                         ifelse(month_numeric %in% 6:8, "Summer",
                                ifelse(month_numeric %in% 3:5, "Spring",
                                       "Winter"))))

flights_delayed <- flights_df %>% 
filter(dep_delay>15) %>%
select(airline,month_numeric,day, season, distance,dep_delay)

flights_delayed

ggplot(flights_delayed, aes(season, distance, fill=airline))+ geom_col()+
labs(title = "Flying Distance of airlines by Season")+xlab("Season") +ylab ("Distance")+theme(plot.title =element_text(hjust = 0.5))
```



## Question 8

**Question**:
What are the number of airlines that operate more during the weekdays or weekends based on the month/season?


**Answer**:
In the month of July, most of the flights tend to travel a lot during weekends and in the month of August, flights tend to travel during weekdays. Seasonally, during summer more number of flights travel during weekdays and all the airlines travel the most during weekdays. In conclusion, the Dulles international airport can expect more number of passengers and flights during the weekdays when compared to the weekends.



```{r}
#Segregating days of the week to weekdays and weekends
flights_df<-flights_df%>%
  mutate(weekday_weekend=ifelse(weekday == "Monday"|weekday =="Tuesday"|weekday =="Wednesday"|weekday =="Thursday","Weekday","Weekend"))

#Bar chart for number of flights operating during weekdays/weekends based on the month
ggplot(flights_df, aes(month, fill = weekday_weekend)) + geom_bar()+theme(axis.text.x = element_text(size=8,angle=90))+
labs(title = "Flights operating in Weekdays Vs Weekends")+xlab("Month") +ylab ("Number of flights")+theme(plot.title =element_text(hjust = 0.5))+geom_text(aes(label = ..count..), stat = "count", vjust = 1.5,size=3)

#Bar chart for number of flights operating during weekdays/weekends based on the season
ggplot(flights_df, aes(season, fill = weekday_weekend)) + geom_bar()+theme(axis.text.x = element_text(size=8,angle=90))+
labs(title = "Flights operating in Weekdays Vs Weekends")+xlab("Season") +ylab ("Number of flights")+theme(plot.title =element_text(hjust = 0.5))+geom_text(aes(label = ..count..), stat = "count", vjust = 1.5,size=3)

#Histogram for number of airlines operating during weekdays/weekends
plot_ly(flights_df,x = ~airline, color = ~weekday_weekend)%>%
  layout(title = "Airlines operating in Weekdays Vs Weekends",
         xaxis =list(title = 'Airlines'), 
         yaxis = list(title = 'Number of flights'))

```
## Question 9


**Question**:
How many flights left before 5AM?

**Answer**:
From the given data, it can be defined that June and July had the highest record of early flights.18 flights left before 5AM in June and 14 flights left before 5AM in July. Also, 44 flights left late in the month of July. 



```{r}

flights_df %>%
    group_by(month) %>%
    summarize(early_flight = sum(dep_time<5.00),late_flight=sum(dep_time>23.00))

```

