---
title: "Data Analysis Project"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```



# Student Information

Name: Naman Lahoti
G Number:G01340550




# Introduction

This semester we will be working with a dataset of all domestic outbound flights from Dulles International Airport in 2016.

Airports depend on accurate flight departure and arrival estimates to maintain operations, profitability, customer satisfaction, and compliance with state and federal laws. Flight performance, including departure and arrival delays must be monitored, submitted to the Federal Aviation Agency (FAA) on a regular basis, and minimized to maintain airport operations. **The FAA considered a flight to be delayed if it has an arrival delay of at least 15 minutes**.

The executives at Dulles International Airport have hired you as a Data Science consultant to perform an exploratory data analysis on all domestic flights from 2016 and produce an executive summary of your key insights and recommendations to the executive team.

Before you begin, take a moment to read through the following airline flight terminology to familiarize yourself with the industry: [Airline Flight Terms](https://www.myairlineflight.com)



## Dulles Flights Data

The `flights_df` data frame is loaded below and consists of 33,433 flights from IAD (Dulles International) in 2016. The rows in this data frame represent a single flight with all of the associated features that are displayed in the table below.

**Note**: If you have not installed the `tidyverse` package, please do so by going to the `Packages` tab in the lower right section of RStudio, select the `Install` button and type `tidyverse` into the prompt. If you cannot load the data, then try downloading the latest version of R (at least 4.0). The `readRDS()` function has different behavior in older versions of `R` and may cause loading issues.



```{r}

library(tidyverse)
library(ggplot2)
library(hrbrthemes)
flights_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/dulles_flights.rds'))

```


### Raw Data

```{r}

flights_df

```


# Exploratory Data Analysis


Executives at this company have hired you as a data science consultant to evaluate their flight data and make recommendations on flight operations and strategies for minimizing flight delays.

You must think of **at least 8 relevant questions** that will provide evidence for your recommendations.

The goal of your analysis should be discovering which variables drive the differences between flights that are early/on-time vs. flights that are delayed.

Some of the many questions you can explore include:

-   Are flight delays affected by taxi-out time? Do certain airlines or time of 
    year lead to greater taxi out times (i.e. traffic jams on the runways)?

-   Are certain times of the day or year problematic?

-   Are certain destination or airlines prone to delays?


You must answer each question and provide supporting data summaries with either a summary data frame (using `dplyr`/`tidyr`) or a plot (using `ggplot`) or both.

In total, you must have a minimum of 5 plots and 4 summary data frames for the exploratory data analysis section. Among the plots you produce, you must have at least 4 different types (ex. box plot, bar chart,
histogram, heat map, etc...)

Each question must be answered with supporting evidence from your tables and plots.


## Question 1



**Question**:
Which airlines operate the most from Dulles International Airport and which destination city have more number of flights?

**Answer**:
The data indicates that United airlines has a fleet of 20,653 whose flights travel to different destinations followed by American Airlines having a fleet of 2597 flights. An intriguing observation has been observed where almost out of the 33,433 flights, 61.4% of the flights were operated by United airlines in the year 2016.

Among the 33,433 flights, 4034 flights flew to San Francisco making it as the busiest destination airport city where flights from IAD flew in 2016 and cities like Cleveland, Charlotte, Kansas City, Louisville and St.Louis have only 1 flight operating from the Dulles airport in the entire year. In conclusion, there are more number of flights coming from San Francisco.


To add additional R code chunks for your work, select `Insert` then `R` from the top of this notebook file.

```{r}
count(flights_df,airline)
#Airline carrier frequency
ggplot(flights_df,aes(airline))+geom_bar(position = position_dodge(width = 0.5),fill="light green")+ 
  geom_text(aes(label=..count..),stat="count",vjust = 1.2, colour = "black")+labs(title = "Airline carrier Frequency",position="center")+xlab("Airline")+ylab("Frequency")+theme(plot.title = element_text(hjust = 0.5),axis.text.x = element_text(size=8,angle=90))


#Flights operating to different cities
flights_df%>%
  arrange(desc(dest_airport_city))%>%
  count(dest_airport_city)
```



## Question 2



**Question**:
Is there a certain airline that has a record of number of delays in flight?


**Answer**:
Yes, United airlines has 3006 flights with arrival delay of more than 15 mins as per FAA. But there is one such flight of Delta airlines which has an arrival delay of 914 mins and departure delay of 911 mins, making it the most delayed flight in the entire year.In other words, Alaska airlines have less departure and arrival delay.The majority of flights departing on time arrive late for all the airlines. In conclusion, only by leaving early you can expect to arrive on time, and a reasonable cutoff for this seems to be around five minutes before scheduled takeoff.



```{r}

#Airlines having arrival delay of 15mins are considered as late (FAA)
flights_df%>%
  select(airline,arrival_delay,dep_delay)%>%
  filter(airline =="United"|airline =="American"|airline =="Delta"|airline =="Southwest"|airline =="JetBlue"|airline =="Virgin America"|airline =="Frontier"|airline =="ExpressJet"|airline =="SkyWest"|airline =="Alaska",arrival_delay>15)%>%
  arrange(desc(arrival_delay))%>%
  count(airline)

#Arrival and departure delay of all airlines
flights_delay<-ggplot(flights_df, aes(x= dep_delay, y=arrival_delay,color=airline)) +xlim(-30, 1000) + geom_point()+
  labs(title = "Airlines that have record number of delays")+
  xlab("Departure delay (mins)")+
  ylab("Arrival Delay (mins)")+theme(plot.title = element_text(hjust = 0.5))

ggplotly(flights_delay)

```



## Question 3



**Question**:
Is there any relation between the number of flights and the average_delay?


**Answer**:
From the scatter plot, we can see that there is much greater variation in average delay when there are few flights. For instance from the data, 125 flights having the tail number as N63890 have an average delay of -11.64 mins meaning these flights arrive almost 12 mins early to the airport. In conclusion, we can say that more number of flights have less arrival delays.


```{r}
delays<-flights_df%>%
group_by(tail_num)%>%
  summarize(delay=mean(arrival_delay,na.rm=TRUE),
            n=n())

delays

ggplot(delays,aes(x=n,y=delay))+geom_point(alpha=0.1)+ggtitle("Number of Flights Vs Average Delay")+xlab("Number of flights")+ylab("Average delay (mins)")+theme(plot.title = element_text(hjust = 0.5))

```

## Question 4


**Question**:
 Which destinations have the most airlines operating from Dulles airport?


**Answer**:
From the heatmap/tile chart depicted below, we can see that Atlanta has 5 different airlines i.e. United, Delta, Southwest, Frontier and ExpressJet operating from Dulles Airport. We can also see that United airlines travels to all the cities and operate through out the year. Also there are certain places like Charleston, St.Louis, Kansas City, Louisville, Cleveland, Eagle, Charlotte where only United Airlines operate.

In conclusion, United airlines is the most busiest airlines which travels to almost every city and operates each month of the year.



```{r}
flights_df %>%
    group_by(dest_airport_city) %>%
    summarize(airlines = n_distinct(airline)) %>%
    arrange(desc(airlines))
    ggplot(flights_df, aes(x=month, y=dest_airport_city, 
                           fill=airline,width=.7,height=.7))+
    geom_tile()+theme(axis.text.x = element_text( color="#993333", size=8,angle=90),
    axis.text.y = element_text( color="#993333",size=8))+
    ggtitle("Number of airlines operating to Destination Airport Cities")+
    xlab("Month")+ylab("Destination airport City")+
    theme(plot.title =element_text(hjust = 0.5))
    
    
```




## Question 5



**Question**:

What percentage of flights are delayed by more than one hour?


**Answer**:
The data indicates that  Frontier airlines have always had a departure delay of more than 1 hour. And during the month of July, 27% of the Frontier airlines have had a departure delay of more than 1 hour (July) which is the highest in the year 2016 when compared to other airlines. During the months of March, April, May, July and August, ExpressJet had no departure delay more than one hour.



```{r}
avg_dep_delay_df<-flights_df%>%
  group_by(month_numeric, airline) %>%
  summarize(avg_dep_delay = mean(dep_delay>60)*100,"%") %>%
  arrange(desc(avg_dep_delay))

avg_dep_delay_df

  ggplot(avg_dep_delay_df,mapping = aes(x = month_numeric, y = avg_dep_delay, color = airline)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = 1:12) +
  xlab("Month") +
  ylab("Average Departure Delay Time (percentage)") +
  ggtitle("Airlines having departure delay time more than 1 hour")+
  theme(plot.title =element_text(hjust = 0.5))
  

```


## Question 6



**Question**:
From which cities do flights arrive late to the Dulles International Airport?


**Answer**:
Flights from cities like Minneapolis-St.Paul,Dallas Fort-Worth and Detroit arrive late when compared to other cities. Flights from these cities have an arrival delay of more than 500 minutes. Whereas there are certain cities like San Francisco, Los Angeles, San Diego where certain flights arrive early even before their scheduled arrival time.



```{r}

#Arrival delay of flights from cities
flights_df %>%
  group_by(dest_airport_name,arrival_delay)%>%
  ggplot( aes(x = dest_airport_city, y = arrival_delay)) +
  geom_boxplot()+theme(axis.text.x = element_text(size=8,angle=90))+
  xlab("City Name") +
  ylab("Arrival delay of the flight \n in minutes") +
  ggtitle("Arrival delay of flights from cities")+
  theme(plot.title =element_text(hjust = 0.5))
```


## Question 7



**Question**:
Is there any certain season of the year where flights travel to distant cities having a departure delay of more than 15mins?


**Answer**:
According to the data, during the summer season, flights travel to distant places and among those flights United Airlines travels the most during each season with a departure delay of more than 15 mins. And during the Fall season, less number of flights travel to long distances. In simpler words, September, October and November are the months where airline carriers travel to less distant places.



```{r}
flights_df<-flights_df %>%
 mutate(season = ifelse(month_numeric %in% 9:11, "Fall",
                         ifelse(month_numeric %in% 6:8, "Summer",
                                ifelse(month_numeric %in% 3:5, "Spring",
                                       "Winter"))))

flights_delayed <- flights_df %>% 
filter(dep_delay>15) %>%
select(airline,month_numeric,day, season, distance,dep_delay)

flights_delayed

ggplot(flights_delayed, aes(season, distance, fill=airline))+ geom_col()+
labs(title = "Flying Distance of airlines by Season")+xlab("Season") +ylab ("Distance")+theme(plot.title =element_text(hjust = 0.5))
```



## Question 8

**Question**:
What are the number of airlines that operate more during the weekdays or weekends based on the month/season?


**Answer**:
In the month of July, most of the flights tend to travel a lot during weekends and in the month of August, flights tend to travel during weekdays. Seasonally, during summer more number of flights travel during weekdays and all the airlines travel the most during weekdays. In conclusion, the Dulles international airport can expect more number of passengers and flights during the weekdays when compared to the weekends.



```{r}
#Segregating days of the week to weekdays and weekends
flights_df<-flights_df%>%
  mutate(weekday_weekend=ifelse(weekday == "Monday"|weekday =="Tuesday"|weekday =="Wednesday"|weekday =="Thursday","Weekday","Weekend"))

#Bar chart for number of flights operating during weekdays/weekends based on the month
ggplot(flights_df, aes(month, fill = weekday_weekend)) + geom_bar()+theme(axis.text.x = element_text(size=8,angle=90))+
labs(title = "Flights operating in Weekdays Vs Weekends")+xlab("Month") +ylab ("Number of flights")+theme(plot.title =element_text(hjust = 0.5))+geom_text(aes(label = ..count..), stat = "count", vjust = 1.5,size=3)

#Bar chart for number of flights operating during weekdays/weekends based on the season
ggplot(flights_df, aes(season, fill = weekday_weekend)) + geom_bar()+theme(axis.text.x = element_text(size=8,angle=90))+
labs(title = "Flights operating in Weekdays Vs Weekends")+xlab("Season") +ylab ("Number of flights")+theme(plot.title =element_text(hjust = 0.5))+geom_text(aes(label = ..count..), stat = "count", vjust = 1.5,size=3)

#Histogram for number of airlines operating during weekdays/weekends
plot_ly(flights_df,x = ~airline, color = ~weekday_weekend)%>%
  layout(title = "Airlines operating in Weekdays Vs Weekends",
         xaxis =list(title = 'Airlines'), 
         yaxis = list(title = 'Number of flights'))

```
## Question 9


**Question**:
How many flights left before 5AM?

**Answer**:
From the given data, it can be defined that June and July had the highest record of early flights.18 flights left before 5AM in June and 14 flights left before 5AM in July. Also, 44 flights left late in the month of July. 



```{r}

flights_df %>%
    group_by(month) %>%
    summarize(early_flight = sum(dep_time<5.00),late_flight=sum(dep_time>23.00))

```



# Summary of Results

Write an executive summary of your overall findings and recommendations to the executives at Dulles Airport. Think of this section as your closing remarks of a presentation, where you summarize your key findings and make recommendations on flight operations and strategies for minimizing flight delays.

Your executive summary must be written in a [professional tone](https://www.universalclass.com/articles/writing/business-writing/appropriate-tone-in-business-communications.htm), with minimal grammatical errors, and should include the following
sections:

1.  An introduction where you explain the business problem and goals of your data analysis

    -   What problem(s) is this company trying to solve? Why are they important
        to their future success?

    -   What was the goal of your analysis? What questions were you 
        trying to answer and why do they matter?


2.  Highlights and key findings from your Exploratory Data Analysis section

    -   What were the interesting findings from your analysis and
        **why are they important for the business**?

    -   This section is meant to **establish the need for your recommendations** 
        in the following section
        

3.  Your recommendations to the company

    -   Each recommendation must be supported by your data analysis results

    -   You must clearly explain **why** you are making each recommendation and which results 
        from your data analysis support this recommendation

    -   You must also describe the potential business impact of your recommendation:

        -   Why is this a good recommendation?

        -   What benefits will the business achieve?
      


## Executive Summary

Please write your executive summary below. If you prefer, you can type your summary in a text editor, such as Microsoft Word, and paste your final text here.

**Introduction**

Problem Statement
Delay in airlines is one of the common problems which make the routine of the airport authorities quite hectic sometimes. In many cases it has been found out that airport authorities are held responsible for mismanagement of the work and failing to take due actions which costs the airlines heftily. 

To solve this problem of flight delays, there is a need to analyze the root cause of the delays, to depict which time of the year/month becomes hectic and which airlines cause the greater number of delays in a year. There are certain questions like “Which airlines cause a delay of more than 1 hour?” give an insight to the authorities to take responsible action and prepare for the same.

**Key Findings**
In the year 2016, 33,433 flights flew from Dulles International airport, out of which 20,653 airlines were of United Airlines. And since United airlines has the highest amount of fleet, it is also observed that 3006 flights of United were delayed during 2016. 

It has also been observed that flights from cities like Minneapolis-St. Paul, Dallas-Fort Worth and Detroit have more arrival delays than other cities. These delays from these cities can be caused due to factors like certain late-arriving aircrafts, maintenance or crew problems, baggage loading and fueling etc., Also, there are fewer flights that have an arrival delay of less than 200 mins.

From the insights, it is defined that a greater number of flights are flown during weekdays in the summer season (June, July, August). Interestingly, during the same time many numbers of flights took off from the airport before 5AM.

From the given data, it was analyzed that Frontier airlines had arrival delay of more than 1 hour when compared to other airlines. During the month of June and July there are a greater number of arrival delays of almost all the airlines. In conclusion, the start of the summer season is considered as critical.

**Recommendations** 
It is recommended that since United Airlines have the highest number of delays caused, the authorities should focus more on such airlines since it can be quite hectic for passengers and flights of other airlines due to such delays. 

It may be a good idea to increase the workforce during the months of June, July, August (Summer season) to avoid any bottlenecks at the last moment. 

Flights from cities like Minneapolis-St. Paul, Dallas-Fort Worth and Detroit should be prioritized more to reduce the delay of flights from other cities caused due to these flights. 

Since Frontier airlines had the delays of more than 1 hour in 2016 than any other airlines, the authorities must take necessary actions to know the main cause of such delays and should follow up regarding the same with the airline authorities. 

By solving the above problems using the key findings and recommended solutions, the airport can function smoothly without any bottlenecks and there are also chances that delays by airlines be reduced. 

Since ferrying the passengers and smooth functioning of flights is necessary for any airport, additional information like the airtime of a flight, weather forecast, temperature, passenger count and certain other criteria are very important to gain accurate insights and provide absolute recommendations.








